Topological Graph Theory

### 2019-05-04 Topology/Paper/paper.tex
SO much progress

\documentclass{article}
\input{../../definitions}
\input{../topology}
\input{glossary}
\title{Topological Graph Theory}
\begin{document}
\maketitle
\tableofcontents

@@

\section{Overview}
\subsection{Graphs}
Before we talk about topological graph theory with any level of understanding we must first understand what a \gls{graph} is.

A \gls{graph} is generally defined as a \gls{set} of \glspl{vertex} combined with a \gls{set} of \glspl{edge} between \glspl{vertex}, however here it may be more useful to think about them visually with a simple representation.

Consider first a \gls{set} of points, this may be thought of as just drawing dots on a sheet of paper. Each of these points will be called a \gls{vertex}. Now we may start drawing lines between \glspl{vertex}. Lines may cross over each other and need not be straight. There is no requirement that all \glspl{vertex} have a line going to it. Each of these lines are called an \gls{edge}. We will simply insist that no \gls{edge} connects two \glspl{vertex} and that we do not have multiple \glspl{edge} between the same pair of \glspl{vertex}.

Once we have drawn this we have a representation of a \gls{graph}. If we were to move the \glspl{vertex} around on the paper but leave them having the same \glspl{edge} (the same \glspl{vertex} are connected to the point as they were before). we would be left with the same \gls{graph}. That is to say, it doesn't mater where we put a \gls{vertex} on our sheet, the \gls{graph} exists independently of the representation we draw.

@@

I first came across this question in the 8\textsuperscript{th} grade, and it was presented to us during geometry class. While undoubtedly an interesting problem, it is quite misleading to try and think of this as a geometry problem. Instead we will try and reduce it to a \gls{graph} problem.

Let us start by thinking of every island as a \gls{vertex} and every bridge as an \gls{edge}. We find the following \gls{graph}.

@@

It is worth noting two things about the above diagram. First that the labels on the \glspl{vertex} and \glspl{edge} are unrelated to the problem, but have been added simply to make referring to parts of the \gls{graph} much easier. Second that whatever the above image depicts, does not fit our definition of a \gls{graph}.

Notice that \gls{edge} $2$ and \gls{edge} $3$ both connect \gls{vertex} $A$ to $B$, as well \glspl{edge} $5$ and $6$ do for $B$ and $C$. This is a strict violation of our definition for a \gls{graph}. The issue of course then comes to what would one call such a beast as this where, presumably, one is able to have as many connections between any pair of \glspl{vertex} and could even have connections from a \gls{vertex} to itself.

I am particularly glad that you're paying enough attention to notice that the diagram does not depict a \gls{graph}. This is what we will refer to as a \gls{multigraph}. It is worth noting that \glspl{graph} are a type of \gls{multigraph}, so anything we show to be true for all \glspl{multigraph}, is also true for all \glspl{graph}.

Now to solve this problem we need to make one simple observation about how we walk. If we are to go to island (or vertex) we must also leave that island, unless it is the last island we arrive on. This means, that with the exception of the island we start on and end on, each island must have an even number of bridges connected to it. On the \gls{multigraph} we would say that we need all \glspl{vertex} but a start and end \gls{vertex} to have an even number of edges. If we look at the \gls{multigraph} above, we have four \glspl{vertex} that have an odd number of \glspl{edge} connected to them.

@@

Consider the following \gls{graph}.

@@

This \gls{graph} is called a \gls{complete} as every \gls{vertex} is connected to every other \gls{vertex} by an \gls{edge}; in fact this \gls{graph} in particular is called $K_4$ as it is the complete four vertex \gls{graph}. We would like to find out if we can draw this above \gls{graph} without having any lines crossing. We can in fact draw this \gls{graph} without any intersections and for any skeptics who may being reading this, the below is $K_4$ without any \glspl{edge} intersecting.

@@

So if this \gls{graph} can be drawn without intersection, can any \gls{graph} be drawn without intersections? If some can and some can't how do we tell which can be drawn and which can not? The answer to this comes in Kuratowski's theorem, however before we can even state this theorem we need to build up a bit of terminology for graph theory.

@@

A face is defined as a connected space that contains no edges or verticies and itself is bounded by edges and verticies.

If you think back to high-school geometry and cubes you may recall that a each of the corners is a vertex, the lines connecting verticies are edges and the area between the edges are faces. In a graph we have verticies connected by edges and when we draw them there are empty areas enclosed by edges and verticies. For example the following would correspond with a cube
\begin{center}
	\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
	thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

@@

	\path[every node/.style={font=\sffamily\small}]
	(1) edge node {} (2)
	edge node {} (3)
	edge node {} (5)
	(2) edge node {} (4)
	edge node {} (6)
	(3) edge node {} (4)
	edge node {} (7)
	(4) edge node {} (8)
	(5) edge node {} (6)
	edge node {} (7)
	(6) edge node {} (8)
	(7) edge node {} (8);
	\end{tikzpicture}
\end{center}
Now we can imagine that if we look at the cube straight on maybe $ABDC$ would be the face we are looking at. So too we see the area enclosed in $ABDC$ is a face by the definition we gave. The same is true for $ABFE$, $ACGE$, $DBFH$, and $DCGH$, however that leaves us with only five faces on this cube. The last face must logically come from $EFHG$, however the area enclosed by that contains all the other edges and verticies, so it doesn't fit our definition. However we may notice that the infinite space outside $EFHG$ does not contain any verticies, and therefore we get a sixth face.

@@

\subsection{Path}
Now moving onto a more traditional graph property we have the concept of a path. A path is defined as a finite sequence of verticies with the property that each element of the sequence (excluding the last one) has an edge from it to the next element. For example consider the following graph.

@@

	\path[every node/.style={font=\sffamily\small}]
	(a) edge node {} (b)
	edge node {} (c)
	edge node {} (f)
	(b) edge node {} (d)
	edge node {} (g)
	(c) edge node {} (e)
	edge node {} (h)
	(d) edge node {} (i)
	(e) edge node {} (j)
	(f) edge node {} (g)
	edge node {} (h)
	(g) edge node {} (i)
	(h) edge node {} (j)
	(i) edge node {} (j);
	\end{tikzpicture}
\end{center}

Now the sequence $BGF$ is a path as $B$ connects to $G$ and $G$ connects to $F$. The sequence $ABGFACEJHCABA$ is also a path, however $JEDI$ is not a path as $E$ has no edge to $D$. Notice that how we draw the graph has nothing to do with what is and is not a path.

\subsection{Connected}
A graph is said to be connected if for any pair of verticies, $\pair ab$ there is a path from $a$ to $b$. So if we consider the following graphs 
\begin{center}
	\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
	thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

@@

\subsection{Contraction}
This is not a property, but rather an operation or action that we preform on a graph. Let us consider the following \gls{graph}.

@@

We wish to preform a \gls{contraction} on \gls{edge} $R$. To be clear all graph \glspl{contraction} are on \glspl{edge}. So we will make a new \gls{vertex}, $r\cdot s$ which has all the \glspl{edge} of $r$ and all the \glspl{edge} of $s$ except the \gls{edge} we are contracting across. In this case the \gls{edge} we are contracting across is $R$ and thus we get the following \gls{multigraph}.

@@

This can then be reduced into a \gls{graph} again by simply treating $b_r$ and $b_s$ as the same edge. This operation is particularly important as we will prove that if a \gls{graph} is planar, then so is any \gls{graph} or \gls{multigraph} obtained by contracting an \gls{edge}.

\subsection{Subdivision}
%TODO rename unsubdivide
\def\unsubdivide{vertex replacement}
\def\unsubdividepl{vertex replacements}

To define this we first describe, we first have to describe an operation where we take a vertex $v$ that has degree 2, and replace with an edge as follows.

\begin{center}
	$
	\begin{array}{l}
		\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
		thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

		\node[main node] (1) {$v$};
		\node[main node] (2) [left of=1] {$A$};
		\node[main node] (3) [right of=1] {$B$};

		\path[every node/.style={font=\sffamily\small}]
		(1)	edge node {} (2)
		edge node {} (3);
		\end{tikzpicture}
	\end{array}
	$
	\scalebox{2}{$\to$}
	$
	\begin{array}{l}
		\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
		thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

		\path[every node/.style={font=\sffamily\small}]
		(1)	edge node {} (2);
		\end{tikzpicture}
	\end{array}
	$

	or

	$
	\begin{array}{l}
		\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
		thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

		\node[main node] (1) {$v$};
		\node[main node] (2) [left of=1] {$A$};
		\node[main node] (3) [right of=1] {$B$};

		\path[every node/.style={font=\sffamily\small}]
		(1)	edge node {} (2)
		edge node {} (3)
		(2) edge [bend left] node {} (3);
		\end{tikzpicture}
	\end{array}
	$
	\scalebox{2}{$\to$}
	$
	\begin{array}{l}
		\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
		thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

		\path[every node/.style={font=\sffamily\small}]
		(1)	edge node {} (2);
		\end{tikzpicture}
	\end{array}
	$

	\scalebox{.9}{Here $A$ and $B$ may be connected in any way to a larger graph, however only this subgraph will change.}
\end{center}

Now we are going to call this operation \unsubdivide, and using it we define a graph subdivision as follows.

\begin{definition}[graph subdivision]
	A graph $G$ is a subdivision of $H$ if $H$ may be produced by some series of \unsubdividepl\ on $G$.
\end{definition}

\section{Kuratowski's Theorem}
\begin{theorem}[Kuratowski's Theorem]
	A graph $G$ is nonplanar if and only if it contains a subgraph that is a subdivision of $K_{3,3}$ or $K_5$
\end{theorem}

@@

\begin{proof}
	First consider a connected multigraph with no edges, as this is connected we may only have a single vertex. This produces a single and no edges so we find $$v-e+f=1-0+1=2$$ Now that we know that for 0 edges this rule fits, then either the rule ($v-e+f-2$) is true no mater the number of edges, or there is some number at which point this rule breaks, and $v-e+f\not=2$. If there is a number that breaks this rule, then there must be a lowest number (let's call it $k$) that breaks this rule. For any multigraph with $k$ edges, choose any edge $e$ in the graph.
	\begin{itemize}
		\item If $e$ connects a vertex to itself it is a loop and its removal will result in the loss of one edge and one face (each loop creates a face\footnote{Jordan curve theorem}). The resulting multigraph has less than $k$ edges and thus we know that for it \begin{align*}2&=v-e+f \\&= v-(e+1)+(f+1)\end{align*} and as the original graph had one more edge and one more face it too would fit this rule.
		\item If $e$ is not a loop we may perform an edge contraction on it and this will reduce both the number of edges and the number of vertices by one. Again this gives us a multigraph with less than $k$ edges so we know that for it \begin{align*}2&=v-e+f\\&=(v+1)-(e+1)+f\end{align*} and as the original graph had one more edge and one more vertex it to would fit this rule.
	\end{itemize}

@@

\subsection{Some basic lemmas}
This is the point in the paper where it is no longer written to be understood by everyone. %TODO make this better

\begin{lemma}
	If a graph $G$ has a nonplanar subgraph, then $G$ is nonplanar.
\end{lemma}

\begin{lemma}
	If a graph $G$ is a subdivision of a nonplanar graph, then $G$ is nonplanar.
\end{lemma}

@@

\begin{lemma}
	Every minimal nonplanar graph contains a connected minimal nonplanar subgraph.
\end{lemma}
\begin{proof}
	%TODO redo proof
	Let $G$ be a minimal nonplanar graph. If $G$ is connected then $G$ contains itself as a minimal nonplanar subgraph. If $G$ is not connected then it is composed of connected components. When embedding $G$ within the plane, if every component was planar we could embed them into any space homeomorphic to $\mathbb R^2$. For example each component could be embedded in an open disk with radius $1/2$ about a separate point in $\mathbb Z^2$, and this would then be a plane drawing. This means that at least one component in $G$ must be nonplanar. Now every other component in this graph may not have any edges as their removal would still yield a nonplanar graph (which breaks the definition of minimal nonplanar). This means that there is only one nonplanar component, and further if any edge from this component is to be removed then $G$ would be planar. Now by . From this we get that this component is itself a connected minimal nonplanar subgraph of $G$.
\end{proof}

\printnoidxglossary

>>>>

\documentclass{article}
\input{../../definitions}
\input{../topology}
\title{Topological Graph Theory}
\begin{document}
\maketitle
\tableofcontents

@@

\section{Overview}
\subsection{Graphs}
Before we talk about topological graph theory with any level of understanding we must first understand what a graph is.

A graph is generally defined as a set of vertices combined with a set of edges between vertices, however here it may be more useful to think about them visually with a simple representation.

Consider first a set of points, this may be thought of as just drawing dots on a sheet of paper. Each of these points will be called a vertex. Now we may start drawing lines between vertices. Lines may cross over each other and need not be straight. There is no requirement that all vertices have a line going to it. Each of these lines are called an edge. We will simply insist that no edge connects two vertices and that we do not have multiple edges between the same pair of vertices.

Once we have drawn this we have a representation of a graph. If we were to move the vertices around on the paper but leave them having the same edges (the same vertices are connected to the point as they were before). we would be left with the same graph. That is to say, it doesn't mater where we put a vertex on our sheet, the graph exists independently of the representation we draw.

@@

I first came across this question in the 8\textsuperscript{th} grade, and it was presented to us during geometry class. While undoubtedly an interesting problem, it is quite misleading to try and think of this as a geometry problem. Instead we will try and reduce it to a graph problem.

Let us start by thinking of every island as a vertex and every bridge as an edge. We find the following graph.

@@

It is worth noting two things about the above diagram. First that the labels on the vertices and edges are unrelated to the problem, but have been added simply to make referring to parts of the graph much easier. Second that whatever the above image depicts, does not fit our definition of a graph.

Notice that edge $2$ and edge $3$ both connect vertex $A$ to $B$, as well edges $5$ and $6$ do for $B$ and $C$. This is a strict violation of our definition for a graph. The issue of course then comes to what would one call such a beast as this where, presumably, one is able to have as many connections between any pair of vertices and could even have connections from a vertices to itself.

I am particularly glad that you're paying enough attention to notice that the diagram does not depict a graph. This is what we will refer to as a multigraph. It is worth noting that graphs are a type of multigraph, so anything we show to be true for all multigraphs, is also true for all graphs.

Now to solve this problem we need to make one simple observation about how we walk. If we are to go to island (or vertex) we must also leave that island, unless it is the last island we arrive on. This means, that with the exception of the island we start on and end on, each island must have an even number of bridges connected to it. On the multigraph we would say that we need all vertices but a start and end vertex to have an even number of edges. If we look at the multigraph above, we have four vertices that have an odd number of edges connected to them.

@@

Consider the following graph.

@@

This graph is called a complete as every vertex is connected to every other vertex by an edge; in fact this graph in particular is called $K_4$ as it is the complete four vertex graph. We would like to find out if we can draw this above graph without having any lines crossing. We can in fact draw this graph without any intersections and for any skeptics who may being reading this, the below is $K_4$ without any edges intersecting.

@@

So if this graph can be drawn without intersection, can any graph be drawn without intersections? If some can and some can't how do we tell which can be drawn and which can not? The answer to this comes in Kuratowski's theorem, however before we can even state this theorem we need to build up a bit of terminology for graph theory.

@@

A face is defined as a connected space that contains no edges or vertices and itself is bounded by edges and vertices.

If you think back to high-school geometry and cubes you may recall that a each of the corners is a vertex, the lines connecting vertices are edges and the area between the edges are faces. In a graph we have vertices connected by edges and when we draw them there are empty areas enclosed by edges and vertices. For example the following would correspond with a cube
\begin{center}
	\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
	thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

@@

	\path[every node/.style={font=\sffamily\small}]
	(1) edge node {} (2)
	edge node {} (3)
	edge node {} (5)
	(2) edge node {} (4)
	edge node {} (6)
	(3) edge node {} (4)
	edge node {} (7)
	(4) edge node {} (8)
	(5) edge node {} (6)
	edge node {} (7)
	(6) edge node {} (8)
	(7) edge node {} (8);
	\end{tikzpicture}
\end{center}
Now we can imagine that if we look at the cube straight on maybe $ABDC$ would be the face we are looking at. So too we see the area enclosed in $ABDC$ is a face by the definition we gave. The same is true for $ABFE$, $ACGE$, $DBFH$, and $DCGH$, however that leaves us with only five faces on this cube. The last face must logically come from $EFHG$, however the area enclosed by that contains all the other edges and vertices, so it doesn't fit our definition. However we may notice that the infinite space outside $EFHG$ does not contain any vertices, and therefore we get a sixth face.

@@

\subsection{Walks}
Now moving onto a more traditional graph property we have the concept of a walk. A walk can be thought of as if you were standing on some vertex $v_0$ and just kept following edges to neighboring vertices some number of times. We define it as follows.

\begin{definition}[walk]
	A walk is a finite sequence of vertices and edges with the following properties.
	\begin{itemize}
		\item Every odd element in the sequence is a vertex.
		\item Every even element in the sequence is an edge that connects the vertex before and after it.
		\item A walk always begins and ends with a vertex.
	\end{itemize}
\end{definition}

For example consider the following graph.

@@

	\path[every node/.style={font=\sffamily\small}]
	(a) edge node {$a$} (b)
	edge node {$b$} (c)
	edge node {$c$} (f)
	edge node {$o$} (e)
	(b) edge node {$d$} (d)
	edge node {$e$} (g)
	(c) edge node {$f$} (e)
	edge node {$g$} (h)
	(d) edge node {$h$} (i)
	(e) edge node {$i$} (j)
	(f) edge node {$j$} (g)
	edge node {$k$} (h)
	(g) edge node {$l$} (i)
	(h) edge node {$m$} (j)
	(i) edge node {$n$} (j);
	\end{tikzpicture}
\end{center}

Now the sequence $BeGjF$ is a walk as we can think of starting at vertex $B$, walking down $e$ to $G$, then down $j$ to $F$ and  finishing. The sequence $AaBaA$ is also a walk as we could start at $A$, go down edge $a$ to $B$, then back up $a$ to $A$ again. The sequence $CfEi$ however isn't a walk as it doesn't end with a vertex.

\subsubsection{Path}
\begin{definition}[path]
	A path is a walk where a vertex and edge can not appear twice in the sequence.
\end{definition}

You could think of a path as you start at some vertex and start building a path as you talk your walk. You aren't allows to have your path intersect with itself, so you always

So for example with the above graph $BeGjF$ is not only a walk, but also a path. The sequence $AaBaA$, while it is a walk, it however is not a path as both edge $a$ and vertex $A$ are used twice.

\subsubsection{Cycle}
\begin{definition}[cycle]
	A cycle is a walk, of length greater than one, where a vertex and edge can not appear twice in the sequence with the exception of the first and last vertex which must be the same.
\end{definition}

The definition for cycle is quite similar to that of a path, it effectively is a path that loops back to where it started at the end. The stipulation that a cycle must be "of length greater than one" simply means that the walk $A$ (just the vertex $A$ and nothing else) is not a cycle. Examples of a cycle would include $AbCgHkFcA$, $AbCfEoA$, and $FkHmJnIlGjF$. It is important to note that something like $FcAcF$ is not a cycle as it includes the edge $c$ twice.

\subsection{Connected}
A graph is said to be connected if for any pair of vertices, $\pair ab$ there is a walk from $a$ to $b$. So if we consider the following graphs 
\begin{center}
	\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
	thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

@@

%\subsection{Neighborhood}
%Given some graph $G$ with vertex $v$. The neighborhood of $v$ within $G$ is the set of all vertices that share an edge with $v$, not including $v$.\footnote{If $G$ is a multigraph the definition stays the same, however $v$ maybe it's own neighbor if it has an edge with itself or a loop} Given a subgraph $H$ of $G$, the neighborhood of $H$ within $G$ is a new subgraph of $G$ that includes $H$ as well as any vertex in $G$ that is adjacent to a vertex in $H$ and the edge that they share.
%
%Lets look at some examples
%\begin{center}
%	\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
%	thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]
%	
%		\node[main node] (0) {\textcolor{blue}{$a$}};
%		\node[main node] (1) [above right of=0] {\textcolor{red}{$b$}};
%		\node[main node] (2) [right of=1] {$c$};
%		\node[main node] (3) [below right of=2] {$d$};
%		\node[main node] (4) [below of=3] {\textcolor{red}{$e$}};
%		\node[main node] (5) [below left of=4] {$f$};
%		\node[main node] (6) [left of=5] {$g$};
%		\node[main node] (7) [above left of=6] {$h$};
%		
%		\path[every node/.style={font=\sffamily\small}]
%		(0)
%		edge node {} (2)
%		edge node {} (3)
%		edge node {} (4)
%		(1)
%		edge node {} (3)
%		edge node {} (4)
%		edge node {} (5)
%		(2)
%		edge node {} (4)
%		edge node {} (5)
%		(3)
%		edge node {} (7)
%		(4)
%		edge node {} (6)
%		edge node {} (7)
%		(5)
%		edge node {} (6)
%		edge node {} (7)
%		(6)
%		edge node {} (7);
%	\end{tikzpicture}
%\end{center}
%
%First let us consider the neighborhood of just the vertex \textcolor{blue}{$a$}, this would then be all the vertices that it is adjacent to so we get $$\set{c, d, \textcolor{red}{e}}$$ We could also let $A$ be a subgraph containing only the vertex $\textcolor{blue}{a}$ and so it's neighborhood within $G$ will be the graph below.
%\begin{center}
%	\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
%	thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]
%	
%		\node[main node] (0) {\textcolor{blue}{$a$}};
%		\node[main node] (2) [above right of=0] {$c$};
%		\node[main node] (3) [below right of=2] {$d$};
%		\node[main node] (4) [below of=3] {\textcolor{red}{$e$}};
%		
%		\path[every node/.style={font=\sffamily\small}]
%		(0)
%		edge node {} (2)
%		edge node {} (3)
%		edge node {} (4)
%		(2)
%		edge node {} (4);
%	\end{tikzpicture}
%\end{center}
%The difference is that this is actually a subgraph of $G$, whereas the previous is just vertices.
%
%For another example let $H$ be the subgraph with only the \textcolor{red}{$b$}, \textcolor{red}{$e$} and the edge they share. The neighborhood of $H$ within $G$ within $G$ would then be all of $G$ as every vertex is either a neighbor of $b$ or $e$.
%
\subsection{Contraction}
This is not a property, but rather an operation or action that we preform on a graph. Let us consider the following graph.

@@

We wish to preform a contraction on edge $R$. To be clear all graph contractions are on edges. So we will make a new vertex, $r\cdot s$ which has all the edges of $r$ and all the edges of $s$ except the edge we are contracting across. In this case the edge we are contracting across is $R$ and thus we get the following multigraph.

@@

This can then be reduced into a graph again by simply treating $b_r$ and $b_s$ as the same edge. This operation is particularly important as we will prove that if a graph is planar, then so is any graph or multigraph obtained by contracting an edge.

\subsection{Subdivision}
There are two types of subdivision, an edge subdivision and graph subdivision. Edge subdivisions can best be though of as an operation, much like a contraction. We then can define a graph subdivision in terms of the edge version.

\subsubsection{Edge Subdivision}
An edge subdivision is where we take an an edge and add a vertex into the middle of it. If we have an edge $e$ that goes between vertex $a$ and $b$ then to subdivide it we add a vertex $c$, remove edge $e$ and then add an edge between $a$ and $c$ and between $b$ and $c$. This is demonstrated in the diagram below.
\begin{center}
	$
	\begin{array}{l}
		\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
		thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

			\node[main node] (1) {$a$};
			\node[main node] (2) [right of=1] {$b$};

			\path[every node/.style={font=\sffamily\small}]
			(1)	edge node {$e$} (2);
		\end{tikzpicture}
	\end{array}
	$
	\scalebox{2}{$\to$}
	$
	\begin{array}{l}
		\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

		\node[main node] (1) {$a$};
		\node[main node] (3) [right of=1] {$c$};
		\node[main node] (2) [right of=3] {$b$};

		\path[every node/.style={font=\sffamily\small}]
		(1)	edge node {} (3)
		(2) edge node {} (3);
		\end{tikzpicture}
	\end{array}
	$
\end{center}

This can be thought of as a sort of inverse to contractions. To see this simply preform a contraction on either edge of the output graph above and you will return the the original graph.

\subsubsection{Graph Subdivision}
A graph $G$ is said to be a subdivision of $H$ if $G$ can be obtained from $H$ via some series of edge subdivisions. For example consider the two following graphs.
This can be thought of as just adding vertices along an edge. Intuitively this shouldn't actually change the shape of a graph. This not changing the shape of a graph is something that will be touched on later in lemma \ref{subdivision}.

\section{Kuratowski's Theorem}
\begin{theorem}[Kuratowski's Theorem]\label{THEBIGONE}
	A graph $G$ is nonplanar if and only if it contains a subgraph that is a subdivision of $K_{3,3}$ or $K_5$
\end{theorem}

@@

\begin{proof}
	First consider a connected multigraph with no edges, as this is connected we may only have a single vertex. This produces a single and no edges so we find $$v-e+f=1-0+1=2$$ Now that we know that for 0 edges this rule fits, then either the rule ($v-e+f-2$) is true no mater the number of edges, or there is some number at which point this rule breaks, and $v-e+f\not=2$. If there is a number that breaks this rule, then there must be a lowest number (let's call it $k$) that breaks this rule. For any multigraph with $k$ edges, choose any edge $e$ in the graph.
	\begin{itemize}
		\item If $e$ connects a vertex to itself it is a loop and its removal will result in the loss of one edge and one face (each loop creates a face\footnote{this is due to the Jordan curve theorem}). The resulting multigraph has less than $k$ edges and thus we know that for it \begin{align*}2&=v-e+f \\&= v-(e+1)+(f+1)\end{align*} and as the original graph had one more edge and one more face it too would fit this rule.
		\item If $e$ is not a loop we may perform an edge contraction on it and this will reduce both the number of edges and the number of vertices by one. Again this gives us a multigraph with less than $k$ edges so we know that for it \begin{align*}2&=v-e+f\\&=(v+1)-(e+1)+f\end{align*} and as the original graph had one more edge and one more vertex it to would fit this rule.
	\end{itemize}

@@

\subsection{Some basic lemmas}
Up to this point the paper has been writing such that anyone with a basic mathematical competency and interest could understand and work through the content. From this point on the paper moves a bit faster and assumes knowledge of basic mathematical notation and more experience with math.

The purpose of the next couple sections is to build up the mathematical machinery that we need for our proof of Kuratowski's theorem (theorem \ref{THEBIGONE})

\begin{lemma} \label{subgraph}
	If a graph $G$ has a nonplanar subgraph, then $G$ is nonplanar.
\end{lemma}

\begin{proof}
	If we have a graph $G$ with a nonplanar subgraph $H$ then when drawing $G$ there is a part of the drawing that is $H$. This part of the drawing can not be planar and therefore the drawing of $G$ can not be a plane drawing.
\end{proof}

\begin{lemma} \label{subdivision}
	If a graph $G$ is a subdivision of a nonplanar graph, then $G$ is nonplanar.
\end{lemma}

\begin{proof}

\end{proof}

\begin{lemma} \label{component}
	If a graph $G$ is nonplanar than at least one of it's connected components is nonplanar.
\end{lemma}

@@

\begin{lemma}
	Every minimal nonplanar graph contains a connected minimal nonplanar subgraph
\end{lemma}

\begin{proof}
	Let $G$ be a minimal nonplanar graph. We know by lemma \ref{component} that $G$ contains at least one nonplanar component, $C_1$. Additionally for a graph to be nonplanar it must have edges (as it is trivial to embed a graph of finite vertices with no edges in the plane), so $C_1$ has edges. Now assume for the sake of contradiction that there is another connected component in $G$, $C_2\not=C_1$, and that $C_2$ has at least one edge in it. We know that $C_1$ is nonplanar so if in $G$ we remove an edge from $C_2$ then $G$ would still be planar as it contains $C_1$ (by lemma \ref{subgraph}). Therefore by contradiction $G$ contains only one connected component, $C_1$, with any edges.

	Now to finish the proof assume for the sake of contradiction that $C_1$ is not minimally nonplanar. This means there is some edge in $C_1$, $e$, for which $C_1\setminus\set e$ is nonplanar. Therefore if we remove the same edge $e$ in $G$, $G$ still contains a nonplanar graph $C_1\setminus\set e$, and therefore is nonplanar by lemma \ref{subgraph}. Therefore we have a contradiction as we know $G$ to be minimally nonplanar and thus $C_1$ must also be minimally nonplanar.
\end{proof}

\begin{definition}[cut set]
	In a graph $G$, any set of vertices, whose removal would yield a disconnected graph is called a cut set.
\end{definition}

\begin{definition}[$k$ connected]
	A graph $G$, is said to be $k$ connected if it's minimal cut set is of size $k$.
\end{definition}

\subsection{Some larger theorems}

\begin{theorem}[Menger's theorem] %TODO make sure mutally distinct path is well defined and make sure $n$ connected is well defined
	A graph is $n$ connected if and only if there are at least $n$ mutually distinct paths between any two vertices.
\end{theorem}

\begin{theorem}
	Any connected minimal nonplanar graph is at least 3 connected.
\end{theorem}

\subsection{Bridges}
First consider a planar graph $G$ with cycle $C$. When we make a plane drawing of $G$ we find that $C$ in this drawing is a loop that separates the plane into an interior and exterior. The proof for this comes from the Jordan curve theorem, however we won't be getting into the Jordan curve theorem in this paper. Even so, just to make this complete we must state the Jordan curve theorem, define a Jordan curve and show that $C$ is a Jordan curve.
\subsubsection{The Jordan Curve Theorem}
\begin{theorem}[Jordan curve theorem] %TODO cite wikipedia https://en.wikipedia.org/wiki/Jordan_curve_theorem#Definitions_and_the_statement_of_the_Jordan_theorem
	Let $C$ be a Jordan curve in the plane $\mathbb R^2$. Then its complement, $\mathbb R^2 \setminus C$, consists of exactly two connected components. One of these components is bounded (the interior) and the other is unbounded (the exterior), and the curve C is the boundary of each component.
\end{theorem}

\begin{definition}[Jordan curve]
	A Jordan curve is a curve, $f:\ccint01\to \mathbb R^2$, with the following properties:
	\begin{itemize}
		\item $f$ is continuous
		\item $f$ is injective on the interval $\coint 01$
		\item $f(0) = f(1)$
	\end{itemize}
\end{definition}
\begin{lemma}
	For any plane drawing of a graph, $G$, containing cycle $C$, $C$ is a Jordan curve in the drawing.
\end{lemma}
\begin{proof}
	In order to fully rigorously prove this we have to go back and formalize what a plane drawing is. So let us do this now.

	A drawing is a subset, $D$, of some topological space $S$. For plane drawings this space will be $\mathbb R^2$ with it's standard topology.

	Given a graph $G$, with vertex set $V$ and edge set $E$, a drawing $D$ is said to be a drawing of $G$ if and only if the following criteria are met:
	\begin{itemize}
		\item There exists an injective function $v:V\to D$.
		\item If there is an edge $e\in E$ between vertices $a$ and $b$ then there is a continuous injective function $\hat e:\ccint01\to S$ with $\hat e(0) = v(a)$ and $\hat e(1) = v(b)$ and the image of $e$ is contained within $D$.
		\item For any vertex $a\in V$ and edge $e\in E$ and any real number $x\in\ooint 01$, $\hat e(x)\not=v(a)$.
		\item For any two edges $e\not=f$, $\hat e(x) \not= \hat f(y)$ for all $x,y \in\ooint01$.
		\item for all $d\in D$ there exists either a vertex $a\in V$ such that $v(a) = d$ or there exists an edge $e\in E$ such that $d$ is in the image of $\hat e$.
	\end{itemize}

%TODO ask Shaw what he thinks
	Now consider a plane drawing of a graph $G$ with cycle $C$. We may write out $C$ as $v_1e_1v_2e_2\ldots v_ke_kv_1$. We know that each of these $e_i$s have a corresponding curve $\hat e_i$ that starts at the end of the previous ($\hat e_{i-1}$) curve and ends at the beginning of the next ($\hat e_{i+1}$) curve. This means we can construct a new continuous curve simply by stringing each of these together and then resize the domain. What we end with will then be a curve $f:\ccint 01\to\mathbb R^2$ that is continuous. We will also have $f(0) = f(1)$ as $v_1$ as $e_1$ starts at $v_1$ and $e_k$ ends at $v_1$. We will also have $f$ injective on $\coint{0}{1}$ as we know that edges do not overlap each other (this is formalized in the fourth bullet point of our definition for a drawing). This then means that our cycle creates a Jordan curve.
\end{proof}

\subsubsection{Bridges}

Let $G$ be a connected graph with cycle $C$. Let $G'$ be the graph of $G$ when all edges of $C$ are removed. 
\begin{definition}[partial bridge]
	A partial bridge, $B$, of $C$ in $G$ is any connected subgraph of $G'$ that contains at lest one edge and for any two edges $a$ and $b$ in $B$, there is a walk in $B$ that contains both $a$ and $b$ and does not pass through any vertex of $C$ (it may begin or end on with a vertex of $C$, but no internal vertex may be in $C$). 
\end{definition}
\begin{definition}[bridge]
	A partial bridge, $B$, is a bridge of $C$ in $G$ if there does not exist a partial bridge $B'\not=B$ of $C$ in $G$ such that $B\subset B'$.

	Alternatively a bridge is a maximal partial bridge.
\end{definition}

Bridges are the final tool we use to prove Kuratowski's theorem. We will now build up some lemmas, that we will need and might help us to build up a conception of a bridge. For these let $G$ be a connected graph with cycle $C$ and let $G'$ be the subgraph. All bridges and partial bridges will be in reference to $C$.

\begin{lemma}\label{walktopartialbridge}
	Let $W$ be a walk in $G'$ that does not pass through $C$ (it may start and end in $C$) and has at least one edge. Then the graph, $H$, composed of the vertices and edges in $W$ is a partial bridge.
\end{lemma}
\begin{proof}
	First $H$ is a subgraph of $G'$ as $W$ is a walk in $G'$. Second $H$ contains at least one edge as $W$ contains at least one edge. Third $H$ is connected trivially. Finally $W$ is a walk containing all edges of $H$ and not passing through $C$, so $H$ must be a partial bridge.
\end{proof}
